---
title: "Homework 3"
subtitle: "ECON 470, Spring 2025"
author: "Molly Catlin"
format:
  pdf:
    output-file: "catlin_molly_hwk2_s2"
    output-ext:  "pdf"
    header-includes:
      - \usepackage{float}
      - \floatplacement{table}{H}
---

```{r}
#| include: false

if (!require("pacman")) install.packages("pacman")
pacman::p_load(tidyverse, ggplot2, dplyr, lubridate, readr, readxl, hrbrthemes, fixest,
               scales, gganimate, gapminder, gifski, png, tufte, plotly, OECD,
               ggrepel, survey, foreign, devtools, pdftools, kableExtra, modelsummary,
               kableExtra)
```

Here is a link to my repository: {https://github.com/mollyjc02/Homework_3.git}

\newpage 

```{r}
#| include: false
#| eval: true

setwd("C:/Users/Molly/OneDrive - Emory University/Documents/GitHub/Homework_3/submission_1/results")
load("hwk3_workspace.RData")
```


\newpage 

## 1. Present a bar graph showing the proportion of states with a change in their cigarette tax in each year from 1970 to 1985.

```{r} 
#| echo: false
#| label: tax.change.plot

print(tax.change.plot)
```


\newpage 

## 2. Plot on a single graph the average tax (in 2012 dollars) on cigarettes and the average price of a pack of cigarettes from 1970 to 2018.

```{r} 
#| echo: false
#| label: tax.price.plot

print(tax.price.plot)
```


\newpage 

## 3. Identify the 5 states with the highest increases in cigarette prices (in dollars) over the time period. Plot the average number of packs sold per capita for those states from 1970 to 2018.

```{r} 
#| echo: false
#| label: top.5.plot 

print(top.5.plot)
```


\newpage 

## 4. Identify the 5 states with the lowest increases in cigarette prices over the time period. Plot the average number of packs sold per capita for those states from 1970 to 2018.

```{r} 
#| echo: false
#| label: bot.5.plot

print(bot.5.plot)
```


\newpage 

## 5. Compare the trends in sales from the 5 states with the highest price increases to those with the lowest price increases.

```{r} 
#| echo: false
#| label: comparison.plot

print(comparison.plot)
```


\newpage 

# 1970-1990

## 6a. Focusing only on the time period from 1970 to 1990, regress log sales on log prices to estimate the price elasticity of demand over that period. Interpret your results.

```{r} 
#| echo: false
#| label: model.a

library(broom)

model.a.summary <- summary(model.a)
model.a.summary
```


\newpage 

## 7a. Regress log sales on log prices using the total (federal and state) cigarette tax (in dollars) as an instrument for log prices. Interpret your results and compare your estimates to those without an instrument. Are they different? If so, why?

```{r} 
#| echo: false
#| label: ivs.a

ivs.a.summary <- summary(ivs.a)
ivs.a.summary
```


\newpage 

## 8a. Show the first stage and reduced-form results from the instrument.

```{r} 
#| echo: false
#| label: first.stage.and.reduced.form-a

fs.summary.a <- summary(first.stage.a)
fs.summary.a

rf.summary.a <- summary(reduced.form.a)
rf.summary.a
```


\newpage 

# 1991-2015 (question 9) 

## 6b.

```{r} 
#| echo: false
#| label: model.b

model.b.summary <- summary(model.b)
model.b.summary
```


\newpage 

## 7b. 

```{r} 
#| echo: false
#| label: ivs.b

ivs.b.summary <- summary(ivs.b)
ivs.b.summary
```


\newpage 

## 8b.

```{r} 
#| echo: false
#| label: first.stage.and.reduced.form-b

fs.summary.b <- summary(first.stage.b)
fs.summary.b

rf.summary.b <- summary(reduced.form.b)
rf.summary.b
```


\newpage 

## 10. Compare your elasticity estimates from 1970-1990 versus those from 1991-2015. Are they different? If so, why? 

```{r} 
#| echo: false
#| label: comparison. table 

rownames(comparison.table) <- "Slope Estimate"
kable(comparison.table, col.names = c("1970-1990", "1991-2015"), 
      caption = "Comparison of Slope Estimates for 1970-1990 and 1991-2015", 
      format = "markdown", align = "c")
```
